Resistive magneto-rotational instability in massive protoplanetary disks

### 2013-12-18 paper/appendix.tex
paper update

@@

>>>>

\section{Relation between $Q$ and the Toomre parameter}\label{q3d2d}
The Toomre parameter defined for razor-thin disks is $Q_\mathrm{2D}\equiv \kappa c_s/\pi G
\Sigma$,  where $\Sigma$ is the total column density. To relate our
self-gravity parameter $Q$ and $Q_\mathrm{2D}$, we replace $c_s$ by
$\overline{c_s}\equiv\int\rho c_s dz/\int\rho dz$, and replace
$\kappa$ by $\Omega$, so that 
\begin{align}
  Q_\mathrm{2D} = 2Qf \frac{\int_0^1
    \rho c_s
    dz}{\left(\int_0^1 \rho dz\right)^2},   
\end{align}
where each term on the right-hand-side is non-dimensionalized. 
We display this relation in Fig. \ref{plot_q3d2d} for the isothermal and
polytropic disks.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/q2d_iso}
  \caption{Relation between the self-gravity parameter $Q$ and the
    classic Toomre parameter $Q_\mathrm{2D}$.   
    \label{plot_q3d2d}}
\end{figure}

@@

\section{Modifications for incompressible flow} 
An incompressible flow satisfies 
\begin{align}
  \frac{\p \rho}{\p t} + \bm{v}\cdot\nabla\rho &= 0 \quad \mathrm{and} \quad
  \nabla\cdot\bm{v} = 0
\end{align}
independently in the equilibrium and perturbed flows. Linearizing the
above we obtain 
\begin{align}
  \frac{\dd\rho}{\rho} &= \frac{\ii }{\sigma}\dd v_z
  \left(\ln\rho\right)^\prime,\\
  0 &= \ii k_x \delta v_x + \dd v_z^\prime. 
\end{align}
The Eulerian density and pressure perturbation does not satisify the
barotropic relation assumed for the unperturbed flow. Instead, the
linearized pressure force is
\begin{align}
  -\frac{\nabla \dd P}{\rho} + \frac{\dd\rho}{\rho^2}\nabla P &=
  -\nabla W + \nabla\ln\rho\left(\frac{c_s^2\dd\rho}{\rho} - W\right),
  \notag\\
  & = -\nabla W + \nabla\ln\rho\left(\frac{\ii c_s^2}{\sigma}\dd v _z \ln\rho^\prime  - W\right)
\end{align}
where $W=\delta P/\rho$ as before. The linearized vertical momentum
equation acquires the second term, which gives the vertical velocity as 
\begin{align}
\dd v_z =
\frac{\ii\sigma\left[\tilde{W}^\prime+\left(\ln\rho\right)^\prime W\right]}{\sigma^2 -
  c_s^2\left(\ln{\rho}\right)^{\prime 2}}. 
\end{align}
Differentiating this and using $\delta v_z^\prime = -\ii k_x \delta
v_x$, we obtain, after non-dimensionalizing, 
\begin{align}
  k_x \dd v_x\left[\sigma^2 -
  f^2 c_s^2\left(\ln{\rho}\right)^{\prime 2}\right] + 
  \sigma\left[\tilde{W}^{\prime\prime} +
    \left(\ln\rho\right)^{\prime\prime}W +
    \left(\ln\rho\right)^{\prime}W^\prime\right]f \notag\\ +
  \frac{\sigma c_s^2 \left[\tilde{W}^\prime+\left(\ln{\rho}\right)^\prime
    W\right]\left(\ln{\rho}\right)^\prime}{\sigma^2 -
  f^2c_s^2\left(\ln{\rho}\right)^{\prime 2}}\left[\left(\ln{c_s^2}\right)^\prime\left(\ln{\rho}\right)^\prime
      + 2\left(\ln{\rho}\right)^{\prime\prime}\right]f^3=0. 
\end{align}
The dimensionless linearized Poisson equation becomes
\begin{align}
  \dd\Phi^{\prime\prime} - k_x^2\dd\Phi +
  \frac{\rho\left(\ln{\rho}\right)^\prime\left[\tilde{W}^\prime+\left(\ln\rho\right)^\prime
      W\right]}{Q \left[\sigma^2 -
  f^2c_s^2\left(\ln{\rho}\right)^{\prime 2}\right]} = 0.  
\end{align}
%boundary condition modifications
The boundary condition for the gravitational potential is also
modified to 
\begin{align}
  \dd\Phi^\prime + k_x\dd\Phi +
  \frac{\rho\left[\tilde{W}^\prime+\left(\ln\rho\right)^\prime
      W\right]}{Q\left[\sigma^2 - f^2c_s^2\left(\ln{\rho}\right)^{\prime 2}\right]}=0.
\end{align}

### 2013-12-18 paper/linear.tex
general backup
updated tex
paper update

\section{Linear problem}\label{linear}
We consider axisymmetric Eulerian perturbations to the above
equilibrium in the form $\real[\dd\rho(z)\exp{\imgi(k_xx + \sigma
    t)}]$ and similarly for other fluid variables. Here, $k_x$ is a constant radial wavenumber and 
$\sigma=-(\omega +\imgi\gamma)$ is a complex 
frequency, where $-\omega$ is the real mode frequency and $\gamma$ is
the growth rate. Hereafter, we surpress the exponential factor, as
well as the real part notation.  We further write 
\begin{align}
  &z=\hat{z}H,\quad k_x =  k_x/H,\quad \sigma = \hat{\sigma}\Omega,
  \quad \delta\bm{v} = \csmid 
  \delta\hat{\bm{v}}, \\ 
  &\delta\bm{B} = B\delta\hat{\bm{B}},\quad
  \delta\rho = \rho\hat{W}/\hat{c}_s^2,\quad \delta\Phi =
  \csmid^2\delta\hat{\Phi}, 
\end{align}
where we have introduced the non-dimensional enthalpy perturbation
$\hat{W}$ and the sound-speed $\hat{c}_s=\csmid^{-1}\sqrt{dP/d\rho}$ can be 
obtained from the equation of state. The background frequencies are
written $S=\hat{S}\Omega$ and $\Omega_z=\hat{\Omega}_z\Omega$ and the 
resistivity as $\eta = \hat{\eta}H^2\Omega$.

We will now drop the $\hat{\phantom{a}}$ notation. Henceforth it is
understood that all variables have been appropriately normalized. The
linearized continuity equation is 
\begin{align}
  \frac{\imgi\sigma}{fc_s^2}W + \imgi k_x \dd v_x
  +\left(\ln\rho\right)^\prime\dd v_z + \dd v_z^\prime = 0,\label{lin_cont}
\end{align}
where $^\prime$ denotes $d/dz$. The linearized equations of motion are
\begin{align}
  &\imgi\sigma \dvx - 2\dvy = f\left[v_A^2\dbx^\prime - \imgi k_x
    \left(\w + v_A^2\dbz
    \right)\right],\\
  &\imgi\sigma\dvy + \frac{\kappa^2}{2}\dvx = fv_A^2\dby^\prime,\\
  & \imgi\sigma\dvz = -f\w^\prime,
\end{align}
where the Alfven speed $v_A =
\left(\beta\rho\right)^{-1/2}$, the effective enthalpy perturbation
$\w = W + \dphi$ and the epicycle frequency $\kappa=\sqrt{2(2-S)}$. 
The linearized induction equation is
\begin{align}
&  \imgi\bar{\sigma}\dbx = f\dvx^\prime +
  \eta\dbx^{\prime\prime}+\eta^\prime\dbx^\prime - \imgi k_x
  \eta^\prime \dbz,\label{induct_x}\\
&\imgi\bar{\sigma}\dby = f\dvy^\prime - S\dbx +
  \eta\dby^{\prime\prime}+\eta^\prime\dby^\prime,\label{induct_y}\\
& \imgi\bar{\sigma}\dbz = -\imgi f k_x \dvx + \eta\dbz^{\prime\prime}, \label{induct_vert}
\end{align}
where $\imgi\bar{\sigma} = \imgi\sigma + \eta k_x^2$, and the
divergence-free condition is $\imgi k_x\dbx + \dbz^\prime=0$. Finally,
the linearized Poisson equation is
\begin{align}
  \dphi^{\prime\prime} - k_x^2\dphi = \frac{\rho}{c_s^2f^2Q}W.  \label{lin_poisson}
\end{align}

@@

Note that $L_{31}\propto k_x$, so for $k_x=0$ 
Eq. \ref{lin1}---\ref{lin2} are decoupled from 
Eq. \ref{lin3}---\ref{lin4}. In this case, a solution is $\dphi=W=0$,
with $\dvx$ and $\dvy$ determined by Eq. \ref{lin1}---\ref{lin2},
which is the incompressible MRI problem. This means that the density
and potential perturbations can only influence the MRI for
$k_x\neq0$. However, SG will still have an effect through the
background equilibrium for $k_x=0$.

@@

\subsection{Diagnostics}
We visualize results in terms of energy densities. We define
\begin{align}
  &E_1 \equiv \frac{|\delta \bm{B}|^2}{2\beta},\\
  &E_2 = \frac{\rho}{\hat{c}_s^2}|\real\left(W\dphi^*\right)|,\\
  &E_3 = \frac{1}{2}\rho|\delta\bm{v}|^2,\\
  &E_4 = \frac{\rho|W|^2}{2\hat{c}_s^2},
\end{align}
as the perturbed magnetic, gravitational, kinetic and thermal
energies, respectively, which are functions of $z$. We
then normalize each by $\mathrm{max}(E_1)$.

>>>>

\section{Linear problem}\label{linear}
We consider axisymmetric Eulerian perturbations to the above
equilibrium in the form $\real[\dd\rho(z)\exp{\imgi(k_xx + \sigma
    t)}]$ and similarly for other fluid variables. Here, $k_x$ is a constant radial wavenumber and 
$\sigma=-(\omega +\imgi\gamma)$ is a complex 
frequency, where $-\omega$ is the real mode frequency and $\gamma$ is
the growth rate. Hereafter, we surpress the exponential factor, as
well as the real part notation.  We further write 
\begin{align}
  &z=\hat{z}H,\quad k_x =  \hat{k}_x/H,\quad \sigma = \hat{\sigma}\Omega,
  \quad \delta\bm{v} = \csmid 
  \delta\hat{\bm{v}}, \\ 
  &\delta\bm{B} = B\delta\hat{\bm{B}},\quad
  \delta\rho = \rho\hat{W}/\hat{c}_s^2,\quad \delta\Phi =
  \csmid^2\delta\hat{\Phi}, 
\end{align}
where we have introduced the non-dimensional enthalpy perturbation
$\hat{W}$ and the sound-speed $\hat{c}_s=\csmid^{-1}\sqrt{dP/d\rho}$ can be 
obtained from the equation of state. The background frequencies are
written $S=\hat{S}\Omega$ and $\Omega_z=\hat{\Omega}_z\Omega$ and the 
resistivity as $\eta = \hat{\eta}H^2\Omega$.

We will now drop the $\hat{\phantom{a}}$ notation. Henceforth it is
understood that all variables have been appropriately normalized
(i.e. $\Omega\to 1$ and $H\to 1$), unless otherwise stated. The
linearized continuity equation is 
\begin{align}
  \frac{\imgi\sigma}{fc_s^2}W + \imgi k_x \dd v_x
  +\left(\ln\rho\right)^\prime\dd v_z + \dd v_z^\prime = 0,\label{lin_cont}
\end{align}
where $^\prime$ denotes $d/dz$. The linearized equations of motion are
\begin{align}
  &\imgi\sigma \dvx - 2\dvy = f\left[v_A^2\dbx^\prime - \imgi k_x
    \left(\w + v_A^2\dbz
    \right)\right],\\
  &\imgi\sigma\dvy + \frac{\kappa^2}{2}\dvx = fv_A^2\dby^\prime,\\
  & \imgi\sigma\dvz = -f\w^\prime,
\end{align}
where the Alfven speed $v_A =
\left(\beta\rho\right)^{-1/2}$, the effective enthalpy perturbation
$\w = W + \dphi$ and the epicycle frequency $\kappa=\sqrt{2(2-S)}$. 
The linearized induction equation is
\begin{align}
&  \imgi\bar{\sigma}\dbx = f\dvx^\prime +
  \eta\dbx^{\prime\prime}+\eta^\prime\dbx^\prime - \imgi k_x
  \eta^\prime \dbz,\label{induct_x}\\
&\imgi\bar{\sigma}\dby = f\dvy^\prime - S\dbx +
  \eta\dby^{\prime\prime}+\eta^\prime\dby^\prime,\label{induct_y}\\
& \imgi\bar{\sigma}\dbz = -\imgi f k_x \dvx + \eta\dbz^{\prime\prime}, \label{induct_vert}
\end{align}
where $\imgi\bar{\sigma} = \imgi\sigma + \eta k_x^2$, and the
divergence-free condition is $\imgi k_x\dbx + \dbz^\prime=0$. Finally,
the linearized Poisson equation is
\begin{align}
  \dphi^{\prime\prime} - k_x^2\dphi = \frac{\rho}{c_s^2f^2Q}W.  \label{lin_poisson}
\end{align}

@@

Note that $L_{13},\,L_{14},\,L_{23},\,L_{24}$ and $L_{31}$ are
proportional to $k_x$. Then for for $k_x=0$ 
Eq. \ref{lin1}---\ref{lin2} are decoupled from 
Eq. \ref{lin3}---\ref{lin4}. In this case, a solution is $\dphi=W=0$,
with $\dvx$ and $\dvy$ determined by $L_{11}\dvx + L_{12}\dvy = 0$ and 
$L_{21}\dvx + L_{22}\dvy=0$, which is the incompressible MRI problem. 
This means that the density and potential perturbations can only
influence the MRI for $k_x\neq0$. Nevertheless, for $k_x=0$,
self-gravity can still have an effect through the background
equilibrium.

@@

\subsection{Diagnostics}
We visualize results in terms of energy densities. We define
\begin{align}
  &E_m \equiv \frac{|\delta \bm{B}|^2}{2\beta},\\
  &E_g = \frac{\rho}{c_s^2}|\real\left(W\dphi^*\right)|,\\
  &E_k = \frac{1}{2}\rho|\delta\bm{v}|^2,\\
  &E_t = \frac{\rho|W|^2}{2c_s^2},
\end{align}
as the perturbed magnetic, gravitational, kinetic and thermal
energies, respectively, which are functions of $z$. The total energy
is $E=E_m+E_g+E_k+E_t$ and we then normalize $E_i$ by
$\mathrm{max}(E)$.

### 2013-12-18 paper/results.tex
paper update

\subsection{Influence of self-gravity on MRI at larger horizontal
  wavenumbers} 
As remarked above, for $k_xH\ll 1$ the most unstable linear mode is
the incompressible MRI with vanishing density and potential
perturbtions. For self-gravity to play a role in the linear
pertubrations one must consider $k_xH \neq 0$, which we consider
here.

>>>>

\subsection{Non-zero radial wavenumbers}  
As remarked above, for $k_xH\ll 1$ the most unstable linear mode is
the incompressible MRI with vanishing density and potential 
perturbtions. For self-gravity to play a role in the linear
pertubrations we consider $k_xH \neq 0$ and also an isothermal
disk. From \cite{mamat10}, we expect GI to set in
for $Q\lesssim 0.2$. 
%We expect modes associated with GI have $k_x\sim 1/H$, for modes
%with smaller $k_x$ are stabilized by rotation and

In Fig. \ref{compare_growth_varQ} we plot the growth rate of 
unstable modes in disks with uniform resistivity and varying $Q$.  
The appearence of a bump at $k_xH\sim1$ for $Q< 0.2$ signifies GI. 
%We checked that the perturbed energy is dominated
%by gravitational energy.   
%Because the most unstable MRI mode occurs at $k_xH=0$ while the most
%unstable GI mode occurs at $k_xH\sim1$, 
We expect a sufficiently massive magnetized disk to undergo both MRI
at long radial lengthscales ($k_xH\ll 1$) and GI at intermediate
radial lengthscales ($k_xH\sim 1$). However, unless the system
parameters are finely tuned, the MRI and GI modes do not grow
at the same rate. For example, at $Q=0.17$ GI is expected but
increasing it slightly to $Q=0.2$, MRI dominates. This is largely due
to promoting GI with decreasing $Q$, although lowering $Q$ also
stabilizes the MRI.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth_varQ}
  \caption{Linear growth rates as a function of horizontal wavenumber
    $k_x$ in a resistive disk with $\Lambda_0=0.1$ and varying
    strengths of self-gravity. 
    \label{compare_growth_varQ}}
\end{figure}

%Strictly speaking, given a set of disk parameters, one is only
%concerned with the most unstable linear mode.

%The plot indicates the peak GI growth rate is
%$\gamma\sim 0.3\Omega$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth_varLam}
  \caption{Linear mode growth rates as a function of horizontal wavenumber
    $k_x$ in a self-gravitating disk with $Q=0.18$ and uniform resitivity 
     as measured by the mid-plane Elsasser number $\Lambda_0$.
    \label{compare_growth_varLam}}
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth]{figures/result_Q0d18_lamda0d2}
  \caption{Perturbation energies for a linear mode in a
    self-gravitating disk ($Q=0.18$) with uniform resistivity (such
    that $\Lambda_0=0.2$). 
    The mangetic
    ($E_m$, solid), gravitational ($E_k$, dotted ), kinetic ($E_k$,
    dashed) and thermal ($E_t$, dash-dot) energies are normalized by
    the maximum of the total energy.  
    \label{result_Q0.18_lambda0.2}}
\end{figure}
